Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-01-16 paper/sheaf_lics.tex
pass on section 6

One of the most direct application of homotopical notion to type
theory is the introduction of homotopy types. 
%
Using the analogy that points in a space correspond to elements of a
type and that paths between two points correspond to 
elements of the corresponding identity type (which defines equality in type theory),  
%
an $n$-type is simply a type
for which equality becomes trivial up to level $n$. 
%
Voevodsky has realized that this notion admits a compact inductive definition
internal to type theory, given by
% Our work is mainly based on the stratification by $\Type_n$~:
\begin{defi}
  $\IsType n$ is defined by induction on $n\geqslant -2$~:
  \begin{itemize}
  \item $\IsType {(-2)} X$ if $X$ is a contractible type, \ie $X$
    is pointed by $c:X$, and every other point in $X$ is connected to $c$.
  \item $\IsType {(n+1)} X = \prod_{x,y:X} \IsType n (x=y).$
  \end{itemize}
  Then, $\Type_n = \sum_{X:\Type} \IsType n X$.
\end{defi}
% For a type $X$, to be in $\Type_n$ means that path spaces of $X$ are
% trivial after $n+1$ iteration.
%
We also define some syntactic sugar for contractible types and mere propositions.

@@

It is then fairly obvious that the canonical function $\hat f : A \to
\im(f)$ is (-1)-connected, thus the following proposition holds
(\cite{hottbook}, Lemma 7.6.4)~:
\begin{prop}
  A map $f:A\to B$ factors through $\im(f)$ as a (-1)-connected
  function followed by a (-1)-truncated function.
\end{prop}

@@

As the definition of sheafification is based on subobject classifiers
and thus on homotopy fibers, we need to take a closer look at the way left
exact modalities preserve homotopy fibers.
%
\begin{prop}
\label{sec:defin-basic-prop}
For any $n$-truncated types $X$ and $Y$,
and any map $f:X \to Y$, the modalisation of fiber of $f$ above any element $y:Y$
is the fiber of $\modal f$ above $\eta_Y y$~:
$$\modal \left(\sum_{x:X}  (f x = y)\right) = \sum_{x:\modal X} (\modal f x = \eta_Y y).$$

@@

The proof is based on the following central result.
\begin{lem}
  Let $X:\Type_p$, $Y:\Type_p\modal$ and $f:X\to Y$. If for all $y:Y$,
  $\modal (\fib f y)$ is contractible, then $\modal X = Y$.
\end{lem}
%
As $\sum_{x:\modal X} (\modal f x = \eta_Y y)$ is modal, we
can define $$\phi:\modal \left(\sum_{x:X}  (f x = y)\right) \to
\sum_{x:\modal X} (\modal f x = \eta_Y y).$$
We just need to check that the image of its fibers by $\modal$ are
contractible.
%
From definition of left exactness, one can deduce the following~:
\begin{lem}
  Let $A,B:\Type_p$. Let $f:A\to B$. If $\modal A$ and $\modal B$ are
  contractible, then so is $\fib f b$ for any $b:B$.
\end{lem}
We can rewrite $\fib \phi {x;p}$ in terms of the fiber of a fonction
$\fib {\eta_X} a \to \fib {\eta_Y} b$.

But we have the lemma~:
\begin{lem}
  Let $A:\Type_p$. Then for any $y:\modal A$, $\modal(\fib {\eta_A}
  y)$ is contractible.
\end{lem}

@@

\begin{defi}
  For any type $A$ and $E$, $\iota : A \to E$ and $F:\Type_{n+1}$, we define
    $$
    \Phi_E^\iota : (E \to F) \to (A \to F) 
    $$
    %
    as the map sending an
    arrow $f:E\to F$ to its restriction the restriction of $f \circ m$.
  % \item if $A \to E$ with $n$-truncated homotopy fibers, $\Phi_E^\chi$ is the
  %   map sending $f~:~E \to f$ to $\restriction f A$.
  % \end{itemize}
\end{defi}

@@

From any left exact modality $\modal_{-1}$ on $\HProp$, we define by
induction on the truncation index $n$, a
left exact modality $\modal_n$ at level $n$ that extends $\modal_{-1}$ in the
sense that the following property holds.
%
% We have to be careful~: if we want to propagate the properties of the
% first modality $\modal_{-1}$, all the higher modalities $\modal_n$ must be compatible
% with $\modal_{-1}$, \ie we want the property
\begin{prop}\label{prop:hprop}
  For any mere proposition $P$ (with $\widehat P$ is $P$ seen as a
  $\Type_n$),  $\modal_n \widehat P = \modal_{-1} P$ and the
  following coherence diagram commutes
  $$\xymatrix{
    P \ar@{=}[r] \ar[d]_{\eta_{-1}} & \widehat P \ar[d]^{\eta_n} \\
    \modal_{-1} P \ar@{=}[r] & \modal_n \widehat P 
  }$$
\end{prop}

@@

\nt{je ne comprends pas ce passage, c'est quoi sheafness ? }
For a given induction step, the sheafification is defined in two
steps:
\begin{enumerate}[(i)]
\item {\em separation~:} In this step, we {\em remove} the lack of
  sheafness from the type we are considering.
\item {\em completion~:} In this step, we add what lacks now to the
  separated type to be a sheaf.
\end{enumerate}

@@

% For the case $n=-1$, one can take any left exact modality on
% $\HProp$. Here we took the $\lnot\lnot$ modality~:
% $$\forall P:\HProp,~\modal_j P = \lnot\lnot P.$$
% One can easily show that it defines a left exact modality.

@@

Let $f,g$ be two maps from $\sum_{e:E} (\chi\,e)$ to $\sum_{x:A}
(Bx)$, equal when precomposed with $\pi_1$.
$$\xymatrix @R=4em @C=4em{
  \sum_{e:E} (\chi\, e) \ar@<-2pt>[r]_{g\circ\pi_1} \ar@<2pt>[r]^{f\circ \pi_1} \ar[d]_{\mathrm{dense}}& \sum_{x :A} (Bx) \\
  E \ar@<-2pt>[ru]_{g} \ar@<2pt>[ru]^{f}&
}$$
We can restrict the previous diagram to 
$$\xymatrix @R=4em @C=5em{
  \sum_{e:E} (\chi\, e) \ar@<-2pt>[r]_{\pi_1\circ g\circ\pi_1} \ar@<2pt>[r]^{\pi_1\circ f\circ \pi_1} \ar[d]_{\mathrm{dense}}& \sum_{x :A} (Bx) \\
  E \ar@<-2pt>[ru]_{\pi_1\circ g} \ar@<2pt>[ru]^{\pi_1\circ f}&
}$$
and as $A$ is separated, $\pi_1\circ f = \pi_2 \circ g$.
For the second components, let $x:E$. Notice that for any $x:E$,
$\sum_{y:E} x = y$ has a dense subobject, $\sum_{y:\sum_{e:E} (\chi\,
  e)} x=y_1$~:

@@

If $T$ is already a separated type, the following lemma alllows us
to build a sheaf~:
\begin{lem}
  Let $T:\Type_{n+1}$ be separated, and $U$ be a sheaf. If $T$ embeds
  in $U$, and is closed in $U$, then $T$ is a sheaf.
\end{lem}

@@

The last thing to prove is the cumulativity of our sheafifications~:
\begin{prop}
  If $T:\Type_n$, then $\modal T = \star \widehat T$ (where $\widehat
  T$ is $T$ seen as a $\Type_{n+1}$.
\end{prop}
It will imply that $\star$ is compatible with $\HProp$ in the sense of proposition~\ref{prop:hprop}.

@@

A Coq formalization is available at
\url{https://github.com/KevinQuirin/sheafification/}.
%
It is based on the homotopy type theory library available at
\url{https://github.com/HoTT/HoTT}
\begin{itemize}
\item Formalisation of subobject classifiers has been done in
\texttt{sub\_object\_classifier.v}, 284 lines.
\item 
Colimits, $p$-pullbacks and \v{C}ech nerves has been formalized in
\texttt{colimits.v, cech\_nerve.v}, 552 lines.
\item 
Reflective subuniverses and modalities are formalized in
\texttt{reflective\_subuniverse.v, modalities.v}, 1053 lines.
\item 
Definition of th reflective subuniverse on $\HProp$ is given in
\texttt{sheaf\_base\_case.v}, 186 lines.
\item 
Section~\ref{sec:sheaves} is formalized in
\texttt{sheaf\_def\_and\_thm.v}, 1029 lines.
\item 
Section~\ref{sec:sheafification} is formalized in
\texttt{sheaf\_induction.v}, 2340 lines.
\end{itemize}
At total, with other files containing technical lemmas, the project
contains 6322 lines.
%\nt{describe how many line for each, be clear about what has not been formalized}

In our formalization, some things are admitted, either because they
are just an adaptation of a proposition already formalized
in~\cite{hottbook}, either because they are true, but very technical~:
\begin{itemize}
\item in \texttt{modalities.v}, an equality is admitted in the proof
  of proposition~\ref{sec:defin-basic-prop}.
\item in \texttt{nat\_lemmas.v}, the proof that $\IsHProp (n\leqslant
  m)$ is admitted.
\item in \texttt{sheaf\_induction.v}, in the proof that
  $C(\mu_T)=(C\id)$ (see lemma~\ref{lem:sepiscolim}), we admit the proof
  of equality of projections.
\item in \texttt{sheaf\_induction.v}, we admit some very technical
  proofs in proofs that $\square$ and $\star$ are modalities.
\item in \texttt{sheaf\_induction.v}, cumulativity is an axiom,
  because of universes issues.
\end{itemize}

Some other lemmas are proved, then admitted. We do it so that Coq
doesn't constrain too much universes, without having to explicitly
annotate universes everywhere in the formalization.

The main issue in this formalization is the gestion of universes in
Coq~\cite{sozeau2014universe}. If Coq can handle cumulativity on all
$\Type$, it is not the case for the sigma type $\Type_n$. This issue
can be solved by adding an axiom of cumulativity for $\Type_n$, but as
it would not have any computational content, it would really
complicates the proofs.

Moreover, even the definition of successive sheafification types is
not allowed~:

\[ \left\{
    \begin{array}{l}
      \modal_{-1} = \lambda T:\HProp,~ \lnot\lnot T \\
      \displaystyle{\modal_{n+1} = \lambda T:\Type_{n+1}},\\
      \quad \displaystyle{\sum_{u:T \to \Type_n^\modal} \modal_{-1} 
      \left\|
      \sum_{a:T} u = (\lambda t,~\modal_n (a=t))
      \right\|}
    \end{array}
  \right.,\]
because Coq asks for every $\modal_n$ to have the same universes
levels. Thus, the induction presented in this paper cannot be
formalized at the moment. We only present the induction step.

Talk about issues in universe levels \cite{sozeau2014universe}. Lake of cumulativity for
$\Type_n$ and impossibility to define by induction the sheafification 
because universe levels must be the same for every $n$.

>>>>

One of the most direct application of homotopical notion to type
theory is the introduction of homotopy types. 
%
Using the analogy that points in a space correspond to elements of a
type and that paths between two points correspond to 
elements of the corresponding identity type (which defines equality in type theory),  
%
an $n$-type is simply a type
for which equality becomes trivial up to level $n$. 
%
Voevodsky has realized that this notion admits a compact inductive definition
internal to type theory, given by
% Our work is mainly based on the stratification by $\Type_n$~:
\begin{defi}
  $\IsType n$ is defined by induction on $n\geqslant -2$:
  \begin{itemize}
  \item $\IsType {(-2)} X$ if $X$ is a contractible type, \ie $X$
    is pointed by $c:X$, and every other point in $X$ is connected to $c$.
  \item $\IsType {(n+1)} X = \prod_{x,y:X} \IsType n (x=y).$
  \end{itemize}
  Then, $\Type_n = \sum_{X:\Type} \IsType n X$.
\end{defi}
% For a type $X$, to be in $\Type_n$ means that path spaces of $X$ are
% trivial after $n+1$ iteration.
%
We also define some syntactic sugar for contractible types and mere propositions.

@@

It is then fairly obvious that the canonical function $\hat f : A \to
\im(f)$ is (-1)-connected, thus the following proposition holds
(\cite{hottbook}, Lemma 7.6.4):
\begin{prop}
  A map $f:A\to B$ factors through $\im(f)$ as a (-1)-connected
  function followed by a (-1)-truncated function.
\end{prop}

@@

As the definition of sheafification is based on subobject classifiers
and thus on homotopy fibers, we need to take a closer look at the way left
exact modalities preserve homotopy fibers.
%
\begin{prop}
\label{sec:defin-basic-prop}
For any $n$-truncated types $X$ and $Y$,
and any map $f:X \to Y$, the modalisation of fiber of $f$ above any element $y:Y$
is the fiber of $\modal f$ above $\eta_Y y$:
$$\modal \left(\sum_{x:X}  (f x = y)\right) = \sum_{x:\modal X} (\modal f x = \eta_Y y).$$

@@

The proof is based on the following central result.
\begin{lem}
  Let $X:\Type_p$, $Y:\Type_p\modal$ and $f:X\to Y$. If for all $y:Y$,
  $\modal (\fib f y)$ is contractible, then $\modal X = Y$.
\end{lem}
%
As $\sum_{x:\modal X} (\modal f x = \eta_Y y)$ is modal, we
can define $$\phi:\modal \left(\sum_{x:X}  (f x = y)\right) \to
\sum_{x:\modal X} (\modal f x = \eta_Y y).$$
We just need to check that the image of its fibers by $\modal$ are
contractible.
%
From definition of left exactness, one can deduce the following:
\begin{lem}
  Let $A,B:\Type_p$. Let $f:A\to B$. If $\modal A$ and $\modal B$ are
  contractible, then so is $\fib f b$ for any $b:B$.
\end{lem}
We can rewrite $\fib \phi {x;p}$ in terms of the fiber of a fonction
$\fib {\eta_X} a \to \fib {\eta_Y} b$.

But we have the lemma:
\begin{lem}
  Let $A:\Type_p$. Then for any $y:\modal A$, $\modal(\fib {\eta_A}
  y)$ is contractible.
\end{lem}

@@

\begin{defi}
  For any type $A$ and $E$, $\iota : A \to E$ and $F:\Type_{n+1}$, we define
    $$
    \Phi_E^\iota : (E \to F) \to (A \to F) 
    $$
    %
    as the map sending an
    arrow $f:E\to F$ to its restriction the restriction of $f \circ m$.
  % \item if $A \to E$ with $n$-truncated homotopy fibers, $\Phi_E^\chi$ is the
  %   map sending $f:~E \to f$ to $\restriction f A$.
  % \end{itemize}
\end{defi}

@@

From any left exact modality $\modal_{-1}$ on $\HProp$, we define by
induction on the truncation index $n$, a
left exact modality $\modal_n$ at level $n$ that extends $\modal_{-1}$ in the
sense that the following property holds.
%
% We have to be careful: if we want to propagate the properties of the
% first modality $\modal_{-1}$, all the higher modalities $\modal_n$ must be compatible
% with $\modal_{-1}$, \ie we want the property
\begin{prop}\label{prop:hprop}
  For any mere proposition $P$ (with $\widehat P$ is $P$ seen as a
  $\Type_n$),  $\modal_n \widehat P = \modal_{-1} P$ and the
  following coherence diagram commutes
  $$\xymatrix{
    P \ar@{=}[r] \ar[d]_{\eta_{-1}} & \widehat P \ar[d]^{\eta_n} \\
    \modal_{-1} P \ar@{=}[r] & \modal_n \widehat P 
  }$$
\end{prop}

@@

\nt{je ne comprends pas ce passage, c'est quoi sheafness ? }
For a given induction step, the sheafification is defined in two
steps:
\begin{enumerate}[(i)]
\item {\em separation:} In this step, we {\em remove} the lack of
  sheafness from the type we are considering.
\item {\em completion:} In this step, we add what lacks now to the
  separated type to be a sheaf.
\end{enumerate}

@@

% For the case $n=-1$, one can take any left exact modality on
% $\HProp$. Here we took the $\lnot\lnot$ modality:
% $$\forall P:\HProp,~\modal_j P = \lnot\lnot P.$$
% One can easily show that it defines a left exact modality.

@@

Let $f,g$ be two maps from $\sum_{e:E} (\chi\,e)$ to $\sum_{x:A}
(Bx)$, equal when precomposed with $\pi_1$.
$$\xymatrix @R=4em @C=4em{
  \sum_{e:E} (\chi\, e) \ar@<-2pt>[r]_{g\circ\pi_1} \ar@<2pt>[r]^{f\circ \pi_1} \ar[d]_{\mathrm{dense}}& \sum_{x :A} (Bx) \\
  E \ar@<-2pt>[ru]_{g} \ar@<2pt>[ru]^{f}&
}$$
We can restrict the previous diagram to 
$$\xymatrix @R=4em @C=5em{
  \sum_{e:E} (\chi\, e) \ar@<-2pt>[r]_{\pi_1\circ g\circ\pi_1} \ar@<2pt>[r]^{\pi_1\circ f\circ \pi_1} \ar[d]_{\mathrm{dense}}& \sum_{x :A} (Bx) \\
  E \ar@<-2pt>[ru]_{\pi_1\circ g} \ar@<2pt>[ru]^{\pi_1\circ f}&
}$$
and as $A$ is separated, $\pi_1\circ f = \pi_2 \circ g$.
For the second components, let $x:E$. Notice that for any $x:E$,
$\sum_{y:E} x = y$ has a dense subobject, $\sum_{y:\sum_{e:E} (\chi\,
  e)} x=y_1$:

@@

If $T$ is already a separated type, the following lemma alllows us
to build a sheaf:
\begin{lem}
  Let $T:\Type_{n+1}$ be separated, and $U$ be a sheaf. If $T$ embeds
  in $U$, and is closed in $U$, then $T$ is a sheaf.
\end{lem}

@@

The last thing to prove is the cumulativity of our sheafifications:
\begin{prop}
  If $T:\Type_n$, then $\modal T = \star \widehat T$ (where $\widehat
  T$ is $T$ seen as a $\Type_{n+1}$.
\end{prop}
It will imply that $\star$ is compatible with $\HProp$ in the sense of proposition~\ref{prop:hprop}.

@@

A Coq formalization is available at
\url{https://github.com/KevinQuirin/sheafification}.
%
It is based on the homotopy type theory library available at
\url{https://github.com/HoTT/HoTT}.

We provide a more detailed insight of where properties have been
formalized, and what is the size of the formalisation for each part:
\begin{itemize} 
\item  Homotopy fibers and subobject classifiers are formalized in
\texttt{sub\_object\_classifier.v} (284 lines).
%
\item Colimits, $p$-pullbacks and \v{C}ech nerves are formalized in
\texttt{colimits.v, cech\_nerve.v} (552 lines).
\item
Reflective subuniverses and modalities are formalized in
\texttt{reflective\_subuniverse.v, modalities.v} (1053 lines).
\item 
%
  The definition of the dense topology as a left exact modality on
  $\HProp$ is given in \texttt{sheaf\_base\_case.v} (186 lines).
\item
Section~\ref{sec:sheaves} is formalized in
\texttt{sheaf\_def\_and\_thm.v} (1029 lines).
\item
Section~\ref{sec:sheafification} is formalized in
\texttt{sheaf\_induction.v} (2340 lines).
\end{itemize}

Overall, with other files containing technical lemmas, the project
contains 6322 lines, and it could be reduced a bit by improving the
way Coq tries to rewrite and apply lemmas automatically. 
%
This constitute a significant amount of work but the part dedicated to
sheaves and sheafification is only 4000, which seems quite reasonable
and is encouraging, because it suggests that homotopy type theory
provides a convenient tool to formalize some part of the theory of
higher topoi.

% In our formalization, some things are admitted, either because they
% are just an adaptation of a proposition already formalized
% in~\cite{hottbook}, either because they are true, but very technical:
% \begin{itemize}
% \item in \texttt{modalities.v}, an equality is admitted in the proof
%   of proposition~\ref{sec:defin-basic-prop}.
% \item in \texttt{nat\_lemmas.v}, the proof that $\IsHProp (n\leqslant
%   m)$ is admitted.
% \item in \texttt{sheaf\_induction.v}, in the proof that
%   $C(\mu_T)=(C\id)$ (see lemma~\ref{lem:sepiscolim}), we admit the proof
%   of equality of projections.
% \item in \texttt{sheaf\_induction.v}, we admit some very technical
%   proofs in proofs that $\square$ and $\star$ are modalities.
% \item in \texttt{sheaf\_induction.v}, cumulativity is an axiom,
%   because of universes issues.
% \end{itemize}

In our formalization of sheafification defined
Section~\ref{sec:sheafification} (done in 
\texttt{sheaf\_induction.v}), five very technical facts required to
show that $\square$ and $\modal_{n+1}$ are modalities have been
admitted. Nevertheless, we are confident that they can be proven and
have been admitted only by lack of time.

More importantly, there are few axioms in the formalization that are
due to an unsatisfactory management of polymorphic universe level in
Coq~\cite{sozeau2014universe}.
%
\nt{Say a word about polymorphic universe level, consistency check,
  and the fact that polymorphic universe level are nevertheless
  mandatory for our formalization}

%
If Coq handles cumulativity on all $\Type$ natively, it is not
the case for the $\Sigma$-type $\Type_n$, which require propositional
resizing. 
%
This issue could be solved by adding an axiom of cumulativity
for $\Type_n$ with an explicit management of universes. 
%
But as it would not have any computational content, it
would really complicates the proofs as the axiom would appear
everywhere cumulativity is needed and it would need explicitly
annotation for universe levels everywhere in the formalization.

We have chosen a more simple but less satisfactory solution. 
%
When the absence of cumulativity rises an universe issue for a lemma,
we have proven the lemma and then admit the exact same lemma, just to
let Coq infer less constrained universe levels.

An other issue with universe polymorphism lies in their management for
recursive definitions. Indeed, the definition of successive sheafification types is
not allowed:
%
\[ \left\{
    \begin{array}{l}
      \modal_{-1} = \lambda T:\HProp,~ \lnot\lnot T \\
      \displaystyle{\modal_{n+1} = \lambda T:\Type_{n+1}},\\
      \quad \displaystyle{\sum_{u:T \to \Type_n^\modal} \modal_{-1} 
      \left\|
      \sum_{a:T} u = (\lambda t,~\modal_n (a=t))
      \right\|}
    \end{array}
  \right.\]
%
because Coq asks for every $\modal_n$ to have the same universes
levels in the recursive definition. 
%
Thus, the induction step presented in this paper has been formalized,
but the complete induction can not be formalized at the moment. 
%
This restriction in our formalization can not be solved without
generalizing universe polymorphism for recursive definition.
%
% In particular, compoatcumulativity is an axiom, because of
% universes issues.
